---
title: "Latest_Analysis"
author: "Alyssa Monda"
date: "8/14/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# knitr chunk options set to prevent
# warnings and messages from being
# shown in your final document
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

##Initial Set up
Load packages and set the seed for reproducibility

```{r}
#Downloaded packages 
library(phyloseq)
library(dada2)
library(ggplot2)
library(biom)
library(devtools)
library(knitr)
library(mixOmics)
library(tidyverse)
library(vegan)
library(mixOmics)
library(readxl)
library(tableone)
library(ReporteRs)
library(magrittr)
library(plyr)
library(RColorBrewer)
library(DESeq2)
library(edgeR)
library(dplyr); packageVersion("dplyr")

#If error in phyloseq load visit this page <https://joey711.github.io/phyloseq/install.html>

set.seed(08152017)
```


##Import Meta Data
Import Meta Data as DataFrame

Inorder to input using this code, ensure the file path matches the path to the file on your computer. The file listed is found in the "Latest" folder of the AG data within the sub-folder containing the meta data. Download it to your computer and rewrite the code. 
```{r}
library(readr)
Meta <- read_delim("~/Desktop/ag-cleaned.txt", #file path name
    "\t", escape_double = FALSE, col_types = cols(AGE_YEARS = col_number(), 
        BMI = col_number(), BMI_CORRECTED = col_number(), `ENA-BASE-COUNT` = col_number(),HEIGHT_CM = col_number(), VIOSCREEN_FIBER = col_number(),
        VIOSCREEN_LOW_FAT_DAIRY_SERVING = col_number(), 
        VIOSCREEN_LYSINE = col_number(), 
        VIOSCREEN_MANNITOL = col_number(), 
        VIOSCREEN_SFA100 = col_double(), #capric acid
        VIOSCREEN_SFA40 = col_double(),#SFA4:0 Butyric Acid
        VIOSCREEN_SFA170 = col_double(), #margaric acid
        VIOSCREEN_VEG5_DAY = col_number(), 
        VIOSCREEN_VEGETABLE_SERVINGS = col_number(), 
        chao1_1k = col_double()), trim_ws = TRUE)
```

##Data Exploration for demographic information 

Here I have summarized only a few of the variables in the dataset. In order to find summaries of other variables replace the variable name in the code. 

###Univariate Exploration
```{r}
#Summary Statistics for continuous variables.  
summary(Meta$AGE_YEARS)
summary(Meta$BMI_CORRECTED) #BMI as listed in variable "BMI" listed has implausible values the corrected version has 8213 missing values and a max BMI of 76.89

#Data Visualization
boxplot(Meta$AGE_YEARS)
hist(Meta$AGE_YEARS)
boxplot(Meta$BMI_CORRECTED)
hist(Meta$BMI_CORRECTED)

#Scatterplot BMI*AGE to observe for any trends
plot(Meta$BMI_CORRECTED, Meta$AGE_YEARS)

#Summary Of Categorical Variables

Categ_BodySite<- summary(as.factor(Meta$BODY_SITE))
Categ_BodySite
barplot(Categ_BodySite)

Categ_Age<- summary(as.factor(Meta$AGE_CAT))
Categ_Age
barplot(Categ_Age)

Categ_BMI<- summary(as.factor(Meta$BMI_CAT))
Categ_BMI
barplot(Categ_BMI)

Categ_Race<- summary(as.factor(Meta$RACE))
Categ_Race
barplot(Categ_Race)

Categ_Sex<- summary(as.factor(Meta$SEX))
Categ_Sex

Categ_Education<- summary(as.factor(Meta$LEVEL_OF_EDUCATION))
Categ_Education
barplot(Categ_Education)

Categ_Region<- summary(as.factor(Meta$COUNTRY_RESIDENCE))
Categ_Region


Categ_Birth_Country<- summary(as.factor(Meta$COUNTRY_OF_BIRTH))
Categ_Birth_Country

Categ_Diet<- summary(as.factor(Meta$DIET_TYPE))
Categ_Diet
barplot(Categ_Diet)

Categ_Pregnant<- summary(as.factor(Meta$PREGNANT))
Categ_Pregnant


Categ_Delivery<- summary(as.factor(Meta$CSECTION))
Categ_Delivery
barplot(Categ_Delivery) #CSECTION=TRUE

Categ_Fed<- summary(as.factor(Meta$FED_AS_INFANT))
Categ_Fed
barplot(Categ_Fed)

Categ_Healthy<- summary(as.factor(Meta$SUBSET_HEALTHY))
Categ_Healthy

Categ_CVD<- summary(as.factor(Meta$CARDIOVASCULAR_DISEASE))
Categ_CVD
barplot(Categ_CVD)

Categ_Cancer<- summary(as.factor(Meta$CANCER))
Categ_Cancer
barplot(Categ_Cancer)

Categ_Diabetes<- summary(as.factor(Meta$DIABETES))
Categ_Diabetes
barplot(Categ_Diabetes)

Categ_Sleep<- summary(as.factor(Meta$SLEEP_DURATION))
Categ_Sleep
barplot(Categ_Sleep)

Categ_Pet<- summary(as.factor(Meta$PETS_OTHER))
Categ_Pet

Categ_Dog<- summary(as.factor(Meta$DOG))
Categ_Dog
barplot(Categ_Dog)

Categ_Exercise<- summary(as.factor(Meta$EXERCISE_FREQUENCY))
Categ_Exercise
barplot(Categ_Exercise)

```

###Breakdown by Age Split 

```{r}

#Split Meta for Under age 3
Meta3 <-subset(Meta, AGE_YEARS<3)
summary(Meta3$AGE_YEARS)

#Under 3 and what they are fed
Categ_Fed3<- summary(as.factor(Meta3$FED_AS_INFANT))
Categ_Fed3
barplot(Categ_Fed3)

#Split Meta for Under age 12
Meta12 <-subset(Meta, AGE_YEARS<12)
summary(Meta12$AGE_YEARS)

#Under 12 with Allergies
Categ_Allergies12<- summary(as.factor(Meta12$SEASONAL_ALLERGIES))
Categ_Allergies12
barplot(Categ_Allergies12)

#Split Meta Over 40
Meta40 <-subset(Meta, AGE_YEARS>40)
summary(Meta40$AGE_YEARS)

#Over 40 and CVD
Categ_CVD40<- summary(as.factor(Meta40$CARDIOVASCULAR_DISEASE))
Categ_CVD40
barplot(Categ_CVD40)

```

##BiVariate Exploration

*Allergies and ________ in participants under age 12*
```{r}
#Seasonal Allergies and Dogs


#Seasonal Allergies and Other Pets


#Seasonal Allergies and Delivery Method


#Seasonal Allergies and Food as Infant

```

###List of all variables available for further analysis
```{r}
names(Meta)
```

##Input OTU/Taxonomy 

Create a phyloseq object to work with inside of the phyloseq package. To do this we will upload the ".biom" file then merge the metadata, OTU, taxonomy, and tree file. This will allow us to explore the diversity of the sample. Again check your file names and paths in order to use this code. (Note: ".biom" file is found in the "OTU" folder of the "latest" data. The "meta" folder appears to be missing the ".biom" file associated with the previously used metadata within the "Meta" folder)

```{r}
##Import OTU file
#Data originally downloaded from <ftp://ftp.microbio.me/AmericanGut/latest>
#Make sure to pull the correct biom file to match the meta data down for analysis. 

# To read in original .biom file use command below 
file_path<- ("~/Desktop/otu_table.biom")
dat <- import_biom(file_path)

#To fix error "input string 1 is invalid in this locale" run command "Sys.setlocale(locale="C")" in console and run command chunk again

#Make Phyloseq values for separate pieces of the biom file 
#Taxonomy
taxonomy <- tax_table(dat)
#OTU
OTU_table<- otu_table(dat)

#MetaData
#Create Phyloseq object for sample data so it can be merged
SampleData<- sample_data(Meta)
SD<-sample_data(Meta)

#Import the tree corresponding to this data ### Will Not work
tree<-read_tree_greengenes("~/Desktop/97_otus.tree")

#Merge OTU, Taxonomy, and metadata into one phyloseq object called "ps" that can be used for analysis
Sample_data <- (SD)
sam_cov <- as.data.frame(Sample_data)
rownames(sam_cov) <- sam_cov$"#SampleID"
sd <- sample_data(sam_cov)
sample_names(sd) #Previous steps fix error in naming of SampleIds
otus <- otu_table(OTU_table, taxa_are_rows = TRUE)
sample_names(otus)
tt <- tax_table(taxonomy)
ps <- phyloseq(otus, sd, tt,tree) #Step to merge OTU,Taxonomy, and metadata 
ps #View object and make sure it has all 3 parts

```

###Summary of OTU
```{r}
#Summary of OTU
ntaxa(ps)
rank_names(ps)
nsamples(ps)
sample_names(ps)[1:5]
otu_table(ps)[1:5, 1:5]
tax_table(ps)[1:5, 1:4]
taxa_names(ps)[1:10]
```


